How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-22 Paper/paper.tex
Replaced those with these

\maketitle
\begin{abstract}
Block-based programming languages like Scratch, Alice and Blockly are becoming increasingly common as introductory languages in programming education. There is substantial research showing that those visual programming environments are suitable for teaching programming concepts. But, what do people do when they use Scratch? In this paper we explore the characteristics of Scratch programs. To that end we have scraped the Scratch public repository and retrieved 250.000 projects. We present an analysis of those projects in three different dimensions. Firstly, we look at the types of blocks used and the size of the programs. We then investigate complexity, used abstractions and programming concepts. Finally we look at coding style and detect \emph{code smells and clones}. We find that programming abstraction concepts like functions are not commonly used. We further investigate the presence of code smells and bad programming practices, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially code duplication.
\end{abstract}

@@

The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random manual sampling we found that in some cases those projects contains only sprites and costumes, but no code, while in other programs they were entirety empty apart from the Scratch cat added by default. Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

@@

In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project. We furthermore measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.

We find that the majority of Scratch projects are small; 75\% of the projects have up to 5 sprites, 12 scripts and 76 lines of code, while one fourth of the projects have up to 12 lines of code. On the other end, 5\% of the projects (11,712) have more than 18 sprites and 4.8\% (11,214) consist of more than 500 lines of code. The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.

@@

To understand the complexity of the Scratch programs in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}. Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

The first method for abstraction that we investigate is the use of procedures. In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of those projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

@@

The use of procedures in projects was further investigated through the procedure calls, summarized in Figure \ref{fig:procedurecalls}. A significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects. Most procedures are called exactly once (62.32\% of them) or twice (14.30\%) and from exactly one script (85.92\% of them). Examining the origin of procedure calls, we observed that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, making them recursive calls. Those recursive procedures are found in 1,052 projects\footnote{list of recursive procedures published \todo{link}}.

@@

Users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}. 
Those blocks are used by 29.57\% of the projects. \texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects. The events are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received. This lack of syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).

@@

In total, in the dataset we found 170,532 scripts cloned across sprites in 59,634 (25.54\% of the total non-empty) projects. 726,316 copies of those scripts were found, making each clone being copied an average of 4.26 times. Figure \ref{fig:clonesprojects} plots the distribution of clones across projects. The majority of projects contain up to two cloned scripts; 7.24\% of the projects contain three or more. Figure \ref{fig:clonescopies} plots the number of copies of the identified clones. It is of interest that 79,378 (46.55\%) of the identified clones are copied three or more times, and even in 585 cases from 411 projects they are copied more than 50 times and up to 974 \footnote{\todo{link}}. Manual inspection of the mostly copied clones revealed that they were scripts used multiple times to cater for different parameter values, sometimes in multiple sprites.

@@

The second smell that we examine is the dead code smell. Already in the previous section, Scratch projects were found to have two types of dead code: defined procedures that are not called (5.06\% of the defined procedures in 2,079 projects) and unmatched broadcast-receive messages (in 7.99\% of the projects). In this section we examine an additional case: scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality. A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects. The majority of those scripts (86.6\%) are scripts missing the starting block. Examining the size of those dead scripts, 72.34\% are composed of a single block. As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of those scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610\footnote{\todo{link}}.

@@

Finally, we examine the long method and the large class smell, considering them in the context of Scratch as large script and large sprite smells respectively. For those two smells we use the number of blocks as the size metric. Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset. We used those numbers to split the dataset and retrieve the top 10\% \todo{felienne, can we justify why 10 percent?} largest scripts and sprites, and in this way we found the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively. Using these thresholds, the number of projects exhibiting the large script smell, i.e., containing at least one script with 18 or more blocks, is 69,521 (29.77\% of the total projects) and the number of projects with the large sprite smell is 31,954 (13.68\%).

@@

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
RQ3: Code clones are found in 25.54\% of the projects, with 46.55\% of the clones copied three or more times, in the same or across sprites. 24.36\% of the projects contain scripts that are never invoked, and thus exhibit the dead code smell. In some cases those scripts are large. The large script smell is found in 29.77\% of the projects and the large sprite in 13.68\%.
	\end{minipage}}

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large scale study on \nAnalyzedPrograms projects we scraped from the Scratch repository. We analyzed those projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We found that functions and conditional loops are not commonly used. We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially dead code and code duplication.

>>>>

\maketitle
\begin{abstract}
Block-based programming languages like Scratch, Alice and Blockly are becoming increasingly common as introductory languages in programming education. There is substantial research showing that these visual programming environments are suitable for teaching programming concepts. But, what do people do when they use Scratch? In this paper we explore the characteristics of Scratch programs. To that end we have scraped the Scratch public repository and retrieved 250.000 projects. We present an analysis of these projects in three different dimensions. Firstly, we look at the types of blocks used and the size of the programs. We then investigate complexity, used abstractions and programming concepts. Finally we look at coding style and detect \emph{code smells and clones}. We find that programming abstraction concepts like functions are not commonly used. We further investigate the presence of code smells and bad programming practices, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially code duplication.
\end{abstract}

@@

The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other programs they were entirety empty apart from the Scratch cat added by default. Since these projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

@@

In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project. We furthermore measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of these size metrics.

We find that the majority of Scratch projects are small; 75\% of the projects have up to 5 sprites, 12 scripts and 76 lines of code, while one fourth of the projects have up to 12 lines of code. On the other end, 5\% of the projects (11,712) have more than 18 sprites and 4.8\% (11,214) consist of more than 500 lines of code. The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code\footnote{The Scratch identifiers of these projects can be found in \todo{link}}.

@@

To understand the complexity of the Scratch programs in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of these projects can be found in \todo{link}}. Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

The first method for abstraction that we investigate is the use of procedures. In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of these projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

@@

The use of procedures in projects was further investigated through the procedure calls, summarized in Figure \ref{fig:procedurecalls}. A significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects. Most procedures are called exactly once (62.32\% of them) or twice (14.30\%) and from exactly one script (85.92\% of them). Examining the origin of procedure calls, we observed that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, making them recursive calls. These recursive procedures are found in 1,052 projects\footnote{list of recursive procedures published \todo{link}}.

@@

Users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}. 
These blocks are used by 29.57\% of the projects. \texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects. The events are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received. This lack of syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).

@@

In total, in the dataset we found 170,532 scripts cloned across sprites in 59,634 (25.54\% of the total non-empty) projects. 726,316 copies of these scripts were found, making each clone being copied an average of 4.26 times. Figure \ref{fig:clonesprojects} plots the distribution of clones across projects. The majority of projects contain up to two cloned scripts; 7.24\% of the projects contain three or more. Figure \ref{fig:clonescopies} plots the number of copies of the identified clones. It is of interest that 79,378 (46.55\%) of the identified clones are copied three or more times, and even in 585 cases from 411 projects they are copied more than 50 times and up to 974 \footnote{\todo{link}}. Manual inspection of the mostly copied clones revealed that they were scripts used multiple times to cater for different parameter values, sometimes in multiple sprites.

@@

The second smell that we examine is the dead code smell. Already in the previous section, Scratch projects were found to have two types of dead code: defined procedures that are not called (5.06\% of the defined procedures in 2,079 projects) and unmatched broadcast-receive messages (in 7.99\% of the projects). In this section we examine an additional case: scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality. A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects. The majority of these scripts (86.6\%) are scripts missing the starting block. Examining the size of these dead scripts, 72.34\% are composed of a single block. As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610\footnote{\todo{link}}.

@@

Finally, we examine the long method and the large class smell, considering them in the context of Scratch as large script and large sprite smells respectively. For these two smells we use the number of blocks as the size metric. Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset. We used these numbers to split the dataset and retrieve the top 10\% \todo{felienne, can we justify why 10 percent?} largest scripts and sprites, and in this way we found the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively. Using these thresholds, the number of projects exhibiting the large script smell, i.e., containing at least one script with 18 or more blocks, is 69,521 (29.77\% of the total projects) and the number of projects with the large sprite smell is 31,954 (13.68\%).

@@

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
RQ3: Code clones are found in 25.54\% of the projects, with 46.55\% of the clones copied three or more times, in the same or across sprites. 24.36\% of the projects contain scripts that are never invoked, and thus exhibit the dead code smell. In some cases these scripts are large. The large script smell is found in 29.77\% of the projects and the large sprite in 13.68\%.
	\end{minipage}}

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large scale study on \nAnalyzedPrograms projects we scraped from the Scratch repository. We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We found that functions and conditional loops are not commonly used. We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially dead code and code duplication.
